Challenges to transboundary fisheries management in North America under climate change

### 2020-04-28 Ecol/Manuscript_Draft.Rmd
Updated Ecol after reviewers

---
title: "Challenges to transboundary fisheries management in North America under climate change"
author: "Juliano Palacios-Abrantes^1^ U. Rashid Sumaila^1^ William W. L. Cheung^1^"
subtitle: "[Response to Reviewers]"
date: "^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada"
output:
  word_document: default
  pdf_document: default
bibliography: References/Oceans_Canada_Biblio.bib
editor_options: 
  chunk_output_type: console
csl: References/harvard.csl
---

```{r setup, eval=T, echo=F, warning=F,message=F, results='hide'}

@@

# Abstract

Climate change is shifting the distribution of fish stocks that straddle between Exclusive Economic Zones (EEZ), challenging transboundary fisheries management. In this paper, we examine the projected sharing of jointly managed transboundary fish stocks between the EEZs of Canada and the United States. We hypothesize that ocean warming will alter the sharing of fish stocks between the two countries, and that such changes will intensify under a high carbon emission scenario impacting the stock-share ratio of different stocks. Firstly, we projected changes in potential catch of 33 transboundary stocks of North America in the $21^{st}$ century based on multiple Earth system models’ simulations and species distribution models, under two climate change scenarios. We then look at the specific cases of the International Pacific Halibut Commission that manages pacific halibut (*Hippoglossus stenolepis*) and a resource sharing **DV arrangement** in the Gulf of Maine for cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*) and yellowtail flounder (*Limanda ferruginea*) **RA to discuss the management consequences of shifts in transboundary stocks**. Results show that, even under a low emission scenario, most transboundary fish stocks sharing ratios, i.e., the proportion of the total catch of a fish stock taken by a given country, will change by 2050 relative to present in the direction as expected from the effects of ocean warming. The overall reduction in catch potential, in addition to the changes in stock-share will further exacerbate trade-offs between changes in species catch potential. Such trade-offs in the Atlantic and Pacific regions will be amplified if a high emission scenario is followed, relative to a low carbon emission scenario. Our paper highlights the challenges that transboundary fisheries management will face as species shift their current distribution under a changing climate.

@@

# INTRODUCTION

<!-- ##### First paragraph of paper summary -->
**RA In 1982, the United Nations Law of the Seas (UNCLOS) formalized the concept of economic exclusive zones (EEZs) creating what we know today as sahred stocks [@UnitedNations:1986tl]. Shared stocks can move freely between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua].** Today, an estimated 347 [@Teh:2015gd] to 1500 [@Caddy:1997ue] fish stocks cross national borders, some of them jointly managed by two or more countries. These stocks are responsible for almost 50% of these countries total fish catches [@Teh:2015gd]. **RA While not a law, UNCLOS incentivizes actions to cooperate on the management of shared stocks [@UnitedNations:1986tl]. Today, game theory is one of the most common approaches used to analyze the management of shared stocks, as often, success depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv].** Management of these stocks' **RA can be convoluted** due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf]. In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks (**Koubrak and VanderZwaag, this Special Feature**), .

@@

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd]. To cope with these changes in ocean biophysic properties, marine species, including shared fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], shared fisheries' delicate governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tunas, have significantly increased in some regions such as the subtropical Atlantic and western Pacific Oceans and are projected to continue [@MonllorHurtado:2017cm; @ErauskinExtramiana:2019fw]. Multiple shared species in North America have been observed to shift in distribution following changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the $21^{st}$ century [@Cheung:2018hz]. As a result, some countries or management jurisdictions may see more shared **RA fisheries** and their catches shifting into their waters while others will stand to lose [@Miller:2013iv; @Pinsky:2018cb]. Nevertheless, management rules for shared stocks (e.g. quota or spatial delimitation) are often determined based on current and/or historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp].

@@

The shifts in distribution of shared fish stocks will impact the economics of their fisheries [@Pinsky:2012kq; @Sumaila:2019gh]**Sumaila et al., This Volume**), and create international disputes between countries [@Munro:2002uf; @Spijkers:2017ij;@Pinsky:2018cb]. Canada and the US share important transboundary stocks of salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) offering a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of transboundary fisheries. These countries have a long history of fisheries cooperation participating in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@Merten:2015tu]. Furthermore, climate-related fluctuations in stocks' distribution have historically created disputes between Canada and the US, increasing international conflict and threatening the health of diverse transboundary stocks [@Munro:2002uf; @CIA:2017uta].

@@

It is expected that climate induced shifts in stocks' distribution **RA will** affect the rules in place that **RA keep** international treaties alive. Therefore, the main objective of this article is to assess the level of exposure that bi-lateral transboundary fisheries treaties between Canada and the US have to climate change through shifts in stock distributions. Specifically, we rely on a species distribution model and scenario planning to project the changes in the distribution of 33 fish stocks jointly managed by Canada and the US. Moreover, we use two specific case studies (the International Pacific Halibut Comission -IPHC- and a **DV fisheries arrangement** for the Gulf of Maine -GoMA-) to explore the potential impacts that climate change will have on selected transboundary **RB stocks** management. Finally, we explore similar situations around the world and identify opportunities to improve the adaptability of transboundary **RB stocks** management to climate change in North America. Despite an overall expectation of species following a poleward shift, **RA important geographic constraints (e.g. Gulf of Alaska representing a latitudinal block) [@Kleisner:2016je], geo-political features (e.g. the localization of Alaska in reference to Canada and the contiguous states)**, and management rules (e.g. quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision-makers on the paths to follow under a changing climate.

# METHODS

#### Study Area and Fisheries

The current study focuses on 33 transboundary **RB stocks** jointly managed by Canada and the US for both the Atlantic and the Pacific coasts. The analysis was carried out within the 200 nautical miles of both nation's EEZ from the Gulf of Mexico to The Labrador Sea in the Atlantic, and from California to Chukchi Sea, in the Pacific coast. The species analyzed (**Table A2.1**) were selected based on five treaties in both the Atlantic and Pacific coast of North America (Table 1). All species under the IPHC, Pacific Salmon Comission, the Pacific Whiting Treaty, and the GoMA arrangement were analyzed as these are bi-lateral agreements. However, in the case of NAFO, a subset of 26 species reported was made based on those fished in regions that fall within the EEZs of Canada and the United States (NAFO subareas 2s, 3s with the exception of 3M and 3N and limited to the EEZ, 4s, 5s and 6 A,B, and C) (NAFO n.d.). Data related to fisheries was gathered from the Sea Around Us from 1951 to 2014  [@Zeller2016].

```{r Treaties_Table, eval = T, echo = F}

# Foot note
fn <- "*These species are also part of NAFO. **This is an arrangement and not an official agreement between parties."

data.table::data.table(Treaty =
                         c("International Pacific Halibut Commission (IPHC)",
                           "Pacific Salmon Commission (PSC)",
                           "Pacific Whiting Treaty (PWT)",
                           "Northwest Atlantic Fisheries Organization (NAFO)",
                           "Gulf of Main's Arrangement (GoM Arrangement**)"),
                       Species =
                         c(1,
                           5,
                           1,
                           26,
                           "3*"),
                       Coast =
                         c("Pacific",
                           "Pacific",
                           "Pacific",
                           "Atlantic",
                           "Atlantic")
) %>% 
  kable("latex",
        caption = "Treaties and number of species included in the current analysis."
        ) %>% 
  add_footnote(fn)
```

We used the IPHC and the Gulf of Maine arrangement (hereafter refered as GoMA) as case studies to discuss the implications that climate chage could have in the management of transboundary stocks. For the IPHC, we used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMA we used NAFO's divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Fig. 1).

@@

##### The International Pacific Halibut Comission

The IPHC was established by Canada and the United States to oversee the management of Pacific halibut (*Hippoglossus stenolepis*) [@IPHC:2014tz]. There are 12 regulatory areas from wich 3AB holds 51.2% of the stock, followed by regions 2ABC and 4ACDE with 23.1% and 20.4%, respectively, and lastly region 4B with only 5.2% of the stock distribution [@IPHC:2018vz]. In terms of management, the IPHC implements a total allowable catch (TAC) based on a yearly sampling of the Convention area in addition to a series of regulations to control fishing effort [@IPHC:2018vz]. The TAC is divided between recreational, subsistence and commercial fishery, with a portion set aside for bycatch of other fisheries [@IPHC:2019wt]. The commercial fishing season starts in March ending around November whith restrictions allowing only set line gear with J-type hooks targeting individuals over 81.3 cm with head on [@IPHC:2019wt].

##### The Gulf of Maine Arrangement

Since 1998 Canada and the US celebrate a "Resource Sharing Understanding" to inform the management of George Bank's Atlantic cod (*Gadus morhua*), and Eastern George Bank's haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Pudden:2007bo; @CIA:2017uta; @TRAC:2016wu]. The GoMA suggests catch-limits based on stocks' distribution produced by quaterly surveys and historical catch [@TRAC:2016wu], and so Canada and the US ultimatley take single management decissions [@Soboil:2006gw]. In terms of manangement, the US has a multi species harvest control with area and season cloasures, mesh sizes, effort control, and mobile gear vessels that use bottom otter trawl gear [@Soboil:2006gw]. In contrast, Canada has a quota system in addition to limited-entery licensig, fleet allocations, and mesh and fish size regulation, among other input controls. Canada inshore vessels fish Cod with longline and gillnet while haddock is mainly caught with bottom otter trawl gear [@Soboil:2006gw].

#### Projecting Future Species Distribution

We used a Dynamic Bioclimatic Envelope Model (DBEM) to project the distribution of species from 2015 to 2100 under two scenarios of climate change [@Cheung:2010dt; @Cheung:2016jd]. The DBEM algorithm integrated ecophysiology and habitat suitability with spatial population dynamics of exploited fishes and invertebrates to project shifts in abundance and potential fisheries catches under climate change. The algorithm predicted species distribution based on depth and latitudinal range, habitat preferences and an index of species association with major habitat types to estimate changes in abundance distribution over a 0.5º x 0.5º grid of the world ocean. For each grid cell and time step, the model then calculated species carrying capacity according to sea surface temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry, as well as the species preferences to these conditions. It then incorporated the intrinsic population growth, settled larvae, and net migration of adults from surrounding cells using an advection-diffusion-reaction equation. Finally, the model also simulated the effects of changes in temperature and oxygen content on growth of individuals [@Cheung:2013gk]. Ultimately, the model simulates spatial and temporal population dynamics, and estimates a proxy of maximum sustainable yield (MSY) by applying fishing at MSY level for each grid cell, hereafter referred as maximum catch potential (MCP).

The DBEM was projected using three earth system models (ESM), the Geophysical Fluid Dynamics Laboratory Earth System Models 2M (GFDL)[^GFDL], the Institute Pierre Simon Laplace Climate Model 5 (IPSL-CM5)[^IPSL], and the Max Planck Institute for Meteorology Earth System Model (MPI)[^MPI]. Each model was downscaled to match the DBEM 0.5' x 0.5' grid using the nearest neighbor method, and in some cases, bilinear interpolation [@Cheung:2016hf]. Finally, we used the model outputs for two scenarios of the Intergovernmental Panel on Climate Change (IPCC)-Representative Concentration Pathways (RCP) 2.6 and RCP 8.5 representing a low greenhouse gas emission (strong mitigation) and a high greenhouse gas emission (week mitigation) scenario, respectively  [@IPCC:2014bz]. To estimate model robustness and capture the structural uncertanty build within ESM models, we averaged the DBEM results for all three models ($\mu±\sigma$) and marked regions where at least one ESM disagree in direction with the rest.

[^GFDL]: More information related to the Geophysical Fluid Dynamics Laboratory Earth System Models 2M can be found at www.gfdl.noaa.gov
[^IPSL]: More information related to the Institute Pierre Simon Laplace Climate Model 5 can be found at www.icmc.ipsl.fr/
[^MPI]: More information related to the Max Planck Institute for Meteorology Earth System Model can be found at www.mpimet.mpg.de/en/science/models/

@@

#### Estimation of Maximum Catch Potential Change

For estimating the percentage change of MCP at the regional scale, we frst aggregated the yearly mean MCP of all species per region ($X_{yr}$) and period:

@@

where *y* is year, *r* is region, *s* is species, *n* is total number of species, and $\hat{MCP}$ is the MCP averaged by the three ESMs. In the case of North America as a whole, and the GoM arrangement, region is defined as the 0.5' x 0.5' grid-cell within the EEZ and the specific NAFO regulatory areas, respectiveley. For the IPHC analysis, region is defined as the Comission's regulatory areas (Fig. 1). We then average the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, we computed the regional percentage change in MCP ($\Delta MCP_r$) as follows:

@@

**RB where $X_{t}$ is the future averaged MCP for each of the two time periods analized in this study and $t_0$ is the present averaged MCP ($\mu$ `r Beg_In`-`r Beg_End`). This way, Equ. 2 shows the percentage change in MCP by mid $21^{st}$ century when $X_T$ = $\mu$ `r Mid_In`-`r Mid_End`, and end of $21^{st}$ century when $X_T$ = $\mu$ `r End_In`-`r End_End`, relative to today ($t_0$). The rationale between choosing these time periods was to provide a relative short-term projection (mid-century) that would be more policy-relevant but also show the long term trend (end of century).**

@@

In addition, we borrowed the concept of “threat point” from game theory **RA defined as the minimum payoff that a player is willing to receive in order to cooperate with other players** (see Sumaila et al., this Special Feature). Thus, we estimate the change in the $\hat{MCP}$ **RA (threat point)** that each country **RA (players)** would have for each species (hereafter referred as stock-share ratio), for both the IPHC and the GoMA. **RA The stock-share ratio can be seen as the proportion of the stock's distribution within the study area that each country has**. For this, we first modify Equ. 1, to estimate the aggregated yearly mean MCP of each species per region. We then average the results by the same previouslly metioned periods (present, mid and end of the $21^{st}$ century). Next, for each species we estimated the stock-share ratio ($\alpha_s$) that each region had during each time period:

@@

# RESULTS

```{r Load_Data, echo = F, eval = T, results='hide', warning=F,message=F}

@@

### North American transboundary **RB stocks** are shifting poleward

```{r Changes_Data_All, eval =F, echo = F, fig.height=8, fig.width=12}

@@

  ggsave("New_plot.png",
         plot = p,
         width = 16,
         height = 8,
         units = "in",
         path = Path)

@@

Shifts in the distribution of 33 jointly managed transboundary stocks of North America show that by 2050 MCP of transboundary stocks will increase in Atlantic-Canadian and Alaskan waters and decrease in Pacific-Canadian and the US contiguous states waters (Fig. 2). Such changes will be more intense under the high emission scenario (Fig. 2-A). Under the low emission scenario, MCP will be reduced in a reduced area and with lower intensity (Fig. 2-B). Impacts of cliamte change are expected to intensify towards the end of the $21^{st}$ century, although with greater uncertainty (**Fig A1.1**).

![Fig 2. Percentage change of MCP of transboundary stocks of North America for mid $21^{st}$ century (2041-2060) **RA relative** to present (2005-2014) under a A) lohighw emission scenario and; B) low emission scenario. Points represent areas where at least one model do not agree in the direction of change](./Figures/Fig2.png)

```{r Changes_Per_Species_F, eval=F, echo=F}

@@

# List of DBEM INDEX within the EEZ
#Data provided by Vicki Lam
EEZIDs_List <- read_excel("~/Documents/UBC/Oceans_Project/Distribution/By Cell/Cell_Data/Updated_EEZList_17June2016.xlsx")
EEZ_CellID <- read_excel("~/Documents/UBC/Oceans_Project/Distribution/By Cell/Cell_Data/EEZ_CellID.xlsx")

colnames(EEZ_CellID) <- c("EEZID","INDEX")

# DBEM Coordinate system
Coor <- fread("~/Documents/UBC/Oceans_Project/Distribution/DBEM/Data/Lon_Lat_DBEM.txt",header = FALSE)
colnames(Coor) <- c("INDEX","Longitude","Latitude")

>>>>

---
title: "Challenges to transboundary fisheries management in North America under climate change"
author: "Juliano Palacios-Abrantes^1^ U. Rashid Sumaila^1^ William W. L. Cheung^1^"
subtitle: "[Response to Reviewers]"
date: "^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada"
output:
 word_document: default
 pdf_document: default
bibliography: ../References/Oceans_Canada_Biblio.bib
editor_options: 
 chunk_output_type: console
csl: ../References/E&S.csl
---

```{r setup_oc, eval = T, echo = F, warning = F, message = F, results = 'hide'}

@@

## Abstract

Climate change is shifting the distribution of fish stocks that straddle between Exclusive Economic Zones (EEZ), challenging transboundary fisheries management. Here, we examine the projected sharing of jointly managed transboundary fish stocks between Canada and the United States. We hypothesize that ocean warming will alter the sharing of fish stocks between the two countries, and that such changes will intensify under a high carbon emission scenario. Firstly, we projected changes in potential catch of 33 transboundary stocks of North America in the $21^{st}$ century based on multiple Earth system models’ simulations and species distribution models, under two climate change scenarios. We then look at the specific cases of the International Pacific Halibut Commission that manages pacific halibut (*Hippoglossus stenolepis*) and a resource sharing arrangement in the Gulf of Maine for cod (*Gadus morhua*), haddock (*Melanogrammus aeglefinus*) and yellowtail flounder (*Limanda ferruginea*) to discuss the management consequences of shifts in transboundary stocks. Results show that, even under a low emission scenario, most transboundary fish stocks sharing ratios, i.e., the proportion of the total catch of a fish stock taken by a given country, will change by 2050 relative to present in the direction as expected from the effects of ocean warming. The overall reduction in catch potential, in addition to the changes in stock-share will further exacerbate trade-offs between changes in species catch potential. Such trade-offs in the Atlantic and Pacific regions will be amplified if a high emission scenario is followed, relative to a low carbon emission scenario. Based on the simulation results, we examine possible solution options to reduce climate risks on transboundary fish stocks and fisheries.

@@

## INTRODUCTION

<!-- ##### First paragraph of paper summary -->
In 1982, the United Nations Law of the Seas (UNCLOS) formalized the concept of economic exclusive zones (EEZs) creating what we know today as shared stocks [@UnitedNations:1986tl], i.e. stocks migrate freely between countries EEZs (known as transboundary stocks) or between EEZs and the high seas (also called straddling stocks) [@Song:2017iua]. Today, an estimated 347 [@Teh:2015gd] to 1500 [@Caddy:1997ue] fish stocks cross national borders, some of them jointly managed by two or more countries. These stocks are responsible for almost 50% of these countries total fish catches [@Teh:2015gd]. While not a law, UNCLOS incentivizes actions to cooperate on the management of shared stocks [@UnitedNations:1986tl]. Game theory is one of the most common approaches used to analyze the management of shared stocks, as often, success depends on effective cooperation between parties [@Munro:2002uf; @Sumaila:2013hv]. Management of these stocks' can be convoluted due to the participation of several fishing “players”, different countries and sometimes jurisdictions within a country, the migration patterns of the stock, and their abundance fluctuation within space and time [@Munro:2002uf], Engler and Saunders this Special Feature). In addition, international treaties might not be prepared to address the effects that climate change will bring to shared fish stocks (Engler and Saunders this Special Feature, Koubrak and VanderZwaag, this Special Feature).

@@

The ocean is getting warmer [@Rheim:2013fv], less oxygenated [@Schmidtko:2017cl], and increasing in acidity [@Ross:2011kd; @IPCC:2019to]. To cope with these changes in ocean biophysic properties, marine species, including shared fish stocks, have been shifting their distribution towards the poles and/or deeper waters [@Poloczanska:2016kk]. As climate change reshapes the ocean's environment worldwide [@Gattuso:2015jz], shared fisheries' delicate governance is threatened as new migration patterns may arise [@Miller:2013iv; @Pinsky:2018cb], historic distribution and abundances might shift [@Cheung:2010dt], and species basic natural traits may modify [@Pauly:2017hpa]. Catches of shared stocks like tunas, have significantly increased in some regions such as the subtropical Atlantic and western Pacific Oceans and are projected to continue [@MonllorHurtado:2017cm; @ErauskinExtramiana:2019fw]. Multiple shared species in North America have been observed to shift in distribution following changes in optimal conditions such as sockeye salmon (*Oncorhynchus nerka*) [@McDaniels:2010vn], Atlantic cod [@Pershing:2015gq], and flounders [@Pinsky:2012kq]. Moreover, these shifts are projected to continue towards the end of the $21^{st}$ century [@Cheung:2018hz]. As a result, some countries or management jurisdictions may see more shared fisheries and their catches shifting into their waters while others will stand to lose [@Miller:2013iv; @Pinsky:2018cb]. Nevertheless, management rules for shared stocks (e.g. quota or spatial delimitation) are often determined based on current and/or historic knowledge of the stock's distribution and do not consider future shifts in distributions [@FredstonHermann:2018kp].

@@

The shifts in distribution of shared fish stocks will impact the economics of their fisheries [@Pinsky:2012kq; @Sumaila:2019gh]Sumaila et al., this Special Feature), and create international disputes between countries [@Munro:2002uf; @Spijkers:2017ij;@Pinsky:2018cb]. Canada and the US share important transboundary stocks of salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) offering a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of transboundary fisheries. These countries have a long history of fisheries cooperation participating in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@Merten:2015tu]. Furthermore, climate-related fluctuations in stocks' distribution have historically created disputes between Canada and the US, increasing international conflict and threatening the health of diverse transboundary stocks [@Munro:2002uf; @CIA:2017uta].

It is expected that climate induced shifts in stocks' distribution will affect the rules in place that keep international treaties alive. Therefore, the main objective of this article is to assess the level of exposure that bi-lateral transboundary fisheries treaties between Canada and the US have to climate change through shifts in stock distributions. Specifically, we rely on a species distribution model and scenario planning to project the changes in the distribution of 33 fish stocks jointly managed by Canada and the US. Moreover, we use two specific case studies (the International Pacific Halibut Commission -IPHC- and a fisheries arrangement for the Gulf of Maine -GoMA-) to explore the potential impacts that climate change will have on selected transboundary stocks management. Finally, we explore similar situations around the world and identify opportunities to improve the adaptability of transboundary stocks management to climate change in North America. Despite an overall expectation of species following a poleward shift, important geographic constraints (e.g. Gulf of Alaska representing a latitudinal block) [@Kleisner:2016je], geo-political features (e.g. the localization of Alaska in reference to Canada and the contiguous states), and management rules (e.g. quota allocations, spatial management rules) may play an important role in the redistribution of benefits. Understanding these stocks shifts will shed a light on future conditions and inform decision-makers on the paths to follow under a changing climate.

## METHODS

### Study Area and Fisheries

The current study focuses on 33 transboundary stocks jointly managed by Canada and the US for both the Atlantic and the Pacific coasts. The analysis was carried out within the 200 nautical miles of both nation's EEZ from the Gulf of Mexico to The Labrador Sea in the Atlantic, and from California to Chukchi Sea, in the Pacific coast. The species analyzed (Table A2.1) were selected based on five treaties in both the Atlantic and Pacific coast of North America (Table 1). All species under the IPHC, Pacific Salmon Comission, the Pacific Whiting Treaty, and the GoMA arrangement were analyzed as these are bi-lateral agreements. However, in the case of Northwest Atlantic Fisheries Organization (NAFO), a subset of 26 species reported was made based on those fished in regions that fall within the EEZs of Canada and the United States (NAFO subareas 2s, 3s with the exception of 3M and 3N and limited to the EEZ, 4s, 5s and 6 A,B, and C) [^NAFO]. Data related to fisheries was gathered from the Sea Around Us from 1951 to 2014 [@Zeller2016].

@@

```{r Treaties_Table, eval = F, echo = F}

# Foot note
fn <- "*These species are also part of NAFO. This is an arrangement and not an official agreement between parties."

data.table::data.table(Treaty =
             c("International Pacific Halibut Commission (IPHC)",
              "Pacific Salmon Commission (PSC)",
              "Pacific Whiting Treaty (PWT)",
              "Northwest Atlantic Fisheries Organization (NAFO)",
              "Gulf of Main's Arrangement (GoM Arrangement)"),
            Species =
             c(1,
              5,
              1,
              26,
              "3*"),
            Coast =
             c("Pacific",
              "Pacific",
              "Pacific",
              "Atlantic",
              "Atlantic")
) %>% 
 kable("latex",
    caption = "Treaties and number of species included in the current analysis."
    ) %>% 
 add_footnote(fn)
```

We used the IPHC and the Gulf of Maine arrangement (hereafter referred as GoMA) as case studies to discuss the implications that climate change could have in the management of transboundary stocks. For the IPHC, we used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMA we used NAFO's divisions 5Y, 5Ze, and 4X within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Fig. 1).

@@

### The International Pacific Halibut Comission

The IPHC was established by Canada and the United States to oversee the management of Pacific halibut (*Hippoglossus stenolepis*) [@IPHC:2014tz]. There are 12 regulatory areas from which 3AB holds 51.2% of the stock, followed by regions 2ABC and 4ACDE with 23.1% and 20.4%, respectively, and lastly region 4B with only 5.2% of the stock distribution [@IPHC:2018vz]. In terms of management, the IPHC implements a total allowable catch (TAC) based on a yearly sampling of the Convention area in addition to a series of regulations to control fishing effort [@IPHC:2018vz]. The TAC is divided between recreational, subsistence and commercial fishery, with a portion set aside for bycatch of other fisheries [@IPHC:2019wt]. The commercial fishing season starts in March ending around November whith restrictions allowing only set line gear with J-type hooks targeting individuals over 81.3 cm of total length [@IPHC:2019wt].

### The Gulf of Maine Arrangement

Since 1998 Canada and the US celebrate a "Resource Sharing Understanding" to inform the management of Eastern George Bank's Atlantic cod (*Gadus morhua*) and haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Pudden:2007bo; @CIA:2017uta; @TRAC:2016wu]. The GoMA suggests catch-limits based on a weighted method were 10% represents the stocks’ historical distribution (from 1967 to 1994) and 90% current distributions produced by quarterly surveys and historical catch [@TRAC:2016wu]. In 2017, the allocation proposed for the GoMA was, Atlantic cod 80% Canada to 20 % U.S, haddock 41% Canada and 59% U.S and yellowtail flounder 31% Canada and 69% U.S (TRAC 2016). However, because this is an unofficial agreement, Canada and the US ultimately take single management decisions (Soboil and Sutinen 2006). In terms of management, the US has a multi species harvest control with area and season cloasures, mesh sizes, effort control, and mobile gear vessels that use bottom otter trawl gear [@Soboil:2006gw]. In contrast, Canada has a quota system in addition to limited-entery licensig, fleet allocations, and mesh and fish size regulation, among other input controls. Canada inshore vessels fish cod with longline and gillnet while haddock is mainly caught with bottom otter trawl gear [@Soboil:2006gw].

### Projecting Future Species Distribution

We used a Dynamic Bioclimatic Envelope Model (DBEM) to project the distribution of species from 2015 to 2100 under two scenarios of climate change [@Cheung:2010dt; @Cheung:2016jd]. The DBEM algorithm integrated ecophysiology and habitat suitability with spatial population dynamics of exploited fishes and invertebrates to project shifts in abundance and potential fisheries catches under climate change. The algorithm predicted species distribution based on depth and latitudinal range, habitat preferences and an index of species association with major habitat types to estimate changes in abundance distribution over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world ocean. For each grid cell and time step, the model then calculated species carrying capacity according to sea surface temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry, as well as the species preferences to these conditions. It then incorporated the intrinsic population growth, settled larvae, and net migration of adults from surrounding cells using an advection-diffusion-reaction equation. Finally, the model also simulated the effects of changes in temperature and oxygen content on growth of individuals [@Cheung:2013gk]. Ultimately, the model simulates spatial and temporal population dynamics, and estimates a proxy of maximum sustainable yield (MSY) by applying fishing at MSY level for each grid cell, hereafter referred as maximum catch potential (MCP).

The DBEM was projected using three earth system models (ESM), the Geophysical Fluid Dynamics Laboratory Earth System Models 2M (GFDL)[^GFDL], the Institute Pierre Simon Laplace Climate Model 5 (IPSL-CM5)[^IPSL], and the Max Planck Institute for Meteorology Earth System Model (MPI)[^MPI]. Each model was downscaled to match the DBEM $0.5^{\circ}$ x $0.5^{\circ}$ grid using the nearest neighbor method, and in some cases, bilinear interpolation [@Cheung:2016hf]. Finally, we used the model outputs for two scenarios of the Intergovernmental Panel on Climate Change (IPCC)-Representative Concentration Pathways (RCP) 2.6 and RCP 8.5 representing a low greenhouse gas emission (strong mitigation) and a high greenhouse gas emission (week mitigation) scenario, respectively [@IPCC:2014bz]. To estimate model robustness and capture the structural uncertanty build within ESM models, we averaged the DBEM results for all three models ($\mu±\sigma$) and marked regions where at least one ESM disagree in direction with the rest.

[^GFDL]: Geophysical Fluid Dynamics Laboratory Earth System Models 2M [online] www.gfdl.noaa.gov
[^IPSL]:Institute Pierre Simon Laplace Climate Model 5 [online] www.icmc.ipsl.fr/
[^MPI]: Max Planck Institute for Meteorology Earth System Model [online] www.mpimet.mpg.de/en/science/models/

@@

### Estimation of Maximum Catch Potential Change

For estimating the percentage change of MCP at the regional scale, we first aggregated the yearly mean MCP of all species per region ($X_{yr}$) and period:

@@

where *y* is year, *r* is region, *s* is species, *n* is total number of species, and $\hat{MCP}$ is the MCP averaged by the three ESMs. In the case of North America as a whole, and the GoM arrangement, region is defined as the $0.5^{\circ}$ x $0.5^{\circ}$ grid-cell within the EEZ and the specific NAFO regulatory areas, respectiveley. For the IPHC analysis, region is defined as the Comission's regulatory areas (Fig. 1). We then average the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, we computed the regional percentage change in MCP ($\Delta MCP_r$) as follows:

@@

 where $X_{t}$ is the future averaged MCP for each of the two time periods analized in this study and $t_0$ is the present averaged MCP ($\mu$ `r Beg_In`-`r Beg_End`). This way, Equ. 2 shows the percentage change in MCP by mid $21^{st}$ century when $X_T$ = $\mu$ `r Mid_In`-`r Mid_End`, and end of $21^{st}$ century when $X_T$ = $\mu$ `r End_In`-`r End_End`, relative to today ($t_0$). The rationale between choosing these time periods was to provide a relative short-term projection (mid-century) that would be more policy-relevant but also show the long term-trend (end of century).

@@

In addition, we borrowed the concept of “threat point” from game theory defined as the minimum payoff that a player is willing to receive in order to cooperate with other players (see Sumaila et al., this Special Feature). Thus, we estimate the change in the $\hat{MCP}$ (threat point) that each country (players) would have for each species (hereafter referred as stock-share ratio), for both the IPHC and the GoMA. The stock-share ratio can be seen as the proportion of the stock's distribution within the study area that each country has. For this, we first modify Equ. 1, to estimate the aggregated yearly mean MCP of each species per region. We then average the results by the same previouslly metioned periods (present, mid and end of the $21^{st}$ century). Next, for each species we estimated the stock-share ratio ($\alpha_s$) that each region had during each time period:

@@

## RESULTS

```{r Load_Data, echo = F, eval = F, results='hide', warning=F,message=F}

@@

### North American transboundary stocks are shifting poleward

```{r Changes_Data_All, eval = F, echo = F, fig.height=8, fig.width=12}

@@

 ggsave("Fig2.png",
     plot = p,
     width = 16,
     height = 8,
     units = "in",
     path = Path)

@@

Shifts in the distribution of 33 jointly managed transboundary stocks of North America by 2050 relative to 2010 show that MCP of transboundary stocks will overall increase in Atlantic-Canadian and Alaskan waters and decrease in Pacific-Canadian and the US contiguous states waters (Fig. 2). Such changes will be more intense under the high emission scenario (Fig. 2-A). For most regions, losses of MCP will be more intense under a high emission scenario (Fig 2-A) than under a low emission scenario (Fig 2-B), however, there will be no apparent change in gain intensity between scenarios. Impacts of cliamte change are expected to intensify towards the end of the $21^{st}$ century, although with greater uncertainty (Fig A1.1).

![Fig 2. Percentage change of MCP of transboundary stocks of North America for mid $21^{st}$ century (2041-2060) relative to present (2005-2014) under a A) high emission scenario and; B) low emission scenario. Points represent areas where at least one model do not agree in the direction of change](./Figures/Fig2.png)

```{r Changes_Per_Species_T, eval = F, echo = F }

@@

#### Mid Century result Table ####
Areas <- c(
 "Canada (East Coast)",
 "Canada (Pacific)",
 "USA (East Coast)",
 "USA (West Coast)",
 "USA (Alaska, Arctic)",
 "USA (Alaska, Subarctic)"
)
